
\chapter{Experiments}\label{ch:experiments}

This chapter reports the experimental evaluation of the proposed runtime. It instantiates the methodology of Section~\ref{sec:design-eval} on two reproducible harnesses: the \texttt{runc-hardened-test} laboratory for profile generation and post-exploitation containment (RQ1 and RQ2), and the \texttt{performance-evaluation} repository for runtime cost comparison against stock \texttt{runc} and gVisor (RQ3). Section~\ref{sec:exp-setup} describes the setup; Section~\ref{sec:exp-results} presents the measured outcomes.

\section{Experimental Setup}\label{sec:exp-setup}

\textbf{Runtime-security laboratory (\texttt{runc-hardened-test}).} The primary security experiment runs on a dedicated Ubuntu~24.04 VM with cgroup~v2 and AppArmor enabled. The laboratory builds two runtime binaries---upstream \texttt{runc-stock} and the proposed \texttt{runc-hardened}---and packages a single OCI bundle with an intentionally vulnerable Flask workload (system monitor plus calendar). Command injection in \texttt{/diag/ping} and \texttt{/diag/dns} is retained on purpose: the experiment does not measure whether remote code execution is possible, but what an attacker can do \emph{after} reaching shell execution inside the container. The protocol is:
\begin{enumerate}
    \item run the workload under stock \texttt{runc} without generated profiles (baseline);
    \item repeat a fixed set of post-exploitation payloads through the injection endpoint;
    \item run \texttt{runc-hardened --security-scan} while exercising only legitimate HTTP paths (UI, logs, ping, DNS, calendar upload);
    \item apply the generated AppArmor, seccomp, and capability artefacts;
    \item replay the same payloads under hardened enforcement.
\end{enumerate}
Reproducer scripts live in the laboratory repository; the host requires root for \texttt{runc}, AppArmor profile loading, and BPF tracing.

\textbf{Performance harness (\texttt{performance-evaluation}).} A separate benchmark host compares stock \texttt{runc}, the hardened binary via containerd, gVisor (\texttt{runsc}), and---optionally---an OCI bundle that already carries generated profiles from the security laboratory (\texttt{prepare-hardened-bundle.sh}). Fallback benchmarks report startup latency, parallel scalability (5/10/50 containers), and sysbench CPU/memory inside containers. Each configuration is measured ten times after one warm-up iteration; the median is reported. Table~\ref{tab:perf-startup} is filled from \texttt{results/*/summary.json} once the benchmark archive is complete.

At the time of writing, the security laboratory fully exercises configurations~(a) and~(c) from Section~\ref{sec:design-eval} (default upstream \texttt{runc} vs.\ scan-then-enforce on the proposed runtime). Configurations~(d) and~(e) (gVisor and Kata) are addressed by the performance harness for RQ3; Kata numbers remain optional future work on the same host.

\section{Experimental Results}\label{sec:exp-results}

\subsection{Scope and threat model}

The runtime-security laboratory answers RQ1 and RQ2 on a single representative workload. The attacker model is application-layer command injection followed by arbitrary shell commands in the container context. Kernel-side escape primitives (CVE-specific runtime bugs, mounted \texttt{docker.sock}, privileged containers) are out of scope, consistent with Section~\ref{sec:design-threat}.

\subsection{Profile generation (RQ1)}

A single \texttt{runc run --security-scan} pass, followed by scripted legitimate traffic, produced the expected artefacts under \texttt{oci-bundle/generated/}: \texttt{seccomp.json} with \texttt{defaultAction=SCMP\_ACT\_ERRNO}, an AppArmor profile learned in complain mode, and \texttt{capable-bpfcc.log} from the cgroup-wide BPF trace. Finalisation embedded the seccomp allow-list and narrowed \texttt{process.capabilities} in \texttt{config.json} without widening privileges relative to the trace.

Three profile-quality findings emerged during the run:

\begin{itemize}
    \item \textbf{Seccomp.} The allow-list covered syscalls exercised by the Flask process, subprocess invocations (\texttt{ping}, \texttt{nslookup}), and file I/O on application paths.
    \item \textbf{AppArmor.} Initial runs missed writable paths such as \texttt{/var/lib/sysmon-cal} because the audit collector matched only \texttt{apparmor=DENIED}, while complain mode logs would-be denials as \texttt{apparmor=ALLOWED}. The collector was updated to accept both forms; unit tests now cover calendar upload writes.
    \item \textbf{Capabilities.} The \texttt{capable-bpfcc} trace did not record \texttt{CAP\_NET\_RAW} required by \texttt{/usr/bin/ping} when the capability is carried as a \emph{file capability} (\texttt{security.capability} xattr) rather than through \texttt{cap\_capable} events. Until the runtime merges file capabilities from executed binaries, the laboratory documents an explicit post-scan adjustment (\texttt{REQUIRED\_CAPS=CAP\_NET\_RAW}) in \texttt{apply.sh}.
\end{itemize}

Table~\ref{tab:profquality-lab} summarises the artefact families for the laboratory workload.

\begin{table}[ht]
\centering
\caption{Profile artefacts produced by \texttt{--security-scan} on the runtime-security lab workload.}
\label{tab:profquality-lab}
\begin{tabular}{ll}
\toprule
\textbf{Artefact} & \textbf{Role} \\
\midrule
\texttt{generated/seccomp.json} & Deny-by-default syscall allow-list (\texttt{oci-seccomp-bpf-hook}) \\
\texttt{generated/apparmor.profile} & MAC rules from complain-mode audit \\
\texttt{generated/capable-bpfcc.log} & Observed \texttt{cap\_capable} events (cgroup-wide) \\
\texttt{config.json} (post-finalise) & Narrowed \texttt{process.capabilities}, embedded seccomp \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Containment effectiveness (RQ2)}

After scan-then-enforce, the same injection payloads used in the baseline run were replayed against the hardened configuration. The vulnerable HTTP endpoints remained exploitable by design; the difference appears in post-exploitation actions. Table~\ref{tab:security-lab} records the qualitative outcome observed on the laboratory VM.

\begin{table}[ht]
\centering
\caption{Post-exploitation containment: stock \texttt{runc} (baseline) vs.\ proposed runtime after scan-then-enforce.}
\label{tab:security-lab}
\begin{tabular}{p{2.6cm}p{3.4cm}cc}
\toprule
\textbf{Class} & \textbf{Example payload} & \textbf{Baseline} & \textbf{Hardened} \\
\midrule
Reconnaissance & \texttt{id; hostname; uname -a} & Succeeds & Partially limited \\
Sensitive read & \texttt{cat /etc/shadow} & Succeeds & Blocked (seccomp / AppArmor) \\
Reverse shell & \texttt{bash -i >\& /dev/tcp/\ldots} & Succeeds & Blocked \\
Mount probe & \texttt{mount -t proc proc /mnt} & Succeeds & Blocked \\
Persistence & \texttt{chmod u+s /bin/bash} & Succeeds & Blocked \\
Legitimate ping & \texttt{/diag/ping?host=127.0.0.1} & OK & OK (requires \texttt{CAP\_NET\_RAW} in apply) \\
Calendar upload & \texttt{POST /calendar/upload} & OK & OK (writable path ownership in apply) \\
\bottomrule
\end{tabular}
\end{table}

Baseline execution left an effective in-container root context after injection. Hardened execution blocked reads of sensitive files, reverse-shell setup, and mount-based probing while preserving intended application behaviour once the documented capability and writable-path adjustments were applied. Evidence includes paired HTTP responses, fragments of \texttt{generated/seccomp.json} and \texttt{apparmor.profile}, AppArmor audit lines from \texttt{dmesg} or \texttt{ausearch}, and the capability set printed by \texttt{apply.sh}.

\subsection{Performance (RQ3)}

Steady-state and startup measurements against stock \texttt{runc}, the hardened binary, gVisor, and---optionally---a bundle with enforcement profiles applied are collected with the \texttt{performance-evaluation} harness. Profile generation is a \emph{scan-time} cost (one BPF recording pass); operators pay \emph{enforcement-time} overhead on every subsequent start.

Table~\ref{tab:perf-startup} will be filled from \texttt{results/*/summary.json} after the benchmark host run. Until then, the columns are placeholders. Qualitatively, prior work~\cite{wang2022,volpert2024} suggests the hardened traditional runtime should remain closer to stock \texttt{runc} than gVisor on syscall-heavy microbenchmarks.

\begin{table}[ht]
\centering
\caption{Container performance (median; fallback benchmark, $N=10$).}
\label{tab:perf-startup}
\begin{tabular}{lrrrr}
\toprule
\textbf{Metric} & \textbf{Stock \texttt{runc}} & \textbf{Hardened binary} & \textbf{Hardened + enforcement} & \textbf{gVisor} \\
\midrule
Startup avg (\texttt{/bin/true}), ms & --- & --- & --- & --- \\
Scale-50 parallel, ms                 & --- & --- & --- & --- \\
sysbench CPU (events/s)               & --- & --- & --- & --- \\
sysbench memory (MiB/s)               & --- & --- & --- & --- \\
\bottomrule
\end{tabular}
\end{table}
