
\chapter{Introduction}

\section{Background}

Containers have become the dominant deployment model for modern applications. Unlike virtual machines, which provide hardware-enforced isolation, containers share the host kernel with every other container on the node, which concentrates a substantial attack surface in a single component~\cite{wang2023}. A single kernel vulnerability can therefore compromise every container co-resident on the host~\cite{cve2019}.

The Open Container Initiative (OCI) standardises container runtimes, and its reference implementation, \texttt{runc}, is intentionally biased toward performance and compatibility rather than confinement~\cite{volpert2024}. Linux already provides the kernel primitives that would close the gap: seccomp filters at the syscall boundary~\cite{lopes2020}, AppArmor or SELinux at the mandatory-access-control layer, and Linux capabilities to split the historical \texttt{root} into smaller privileges. In practice, however, all three are only as strong as the policies operators ship with their workloads, and meaningful policies are tedious and error-prone to author by hand.

The industry has reacted in three directions. Sandboxed runtimes such as gVisor and Kata Containers replace the shared-kernel model with either a user-space kernel or a per-container microVM, trading 20--100\% runtime overhead for stronger isolation~\cite{wang2022,viktorsson2020}. Cluster-side projects such as the Kubernetes Security Profiles Operator (SPO)~\cite{spo2025}, Inspektor Gadget~\cite{inspektorgadget2025}, and Tetragon~\cite{tetragon2025} record container behaviour through eBPF and produce seccomp/AppArmor profiles that can later be pinned to pods. Both directions raise the floor of container security significantly, but both also introduce non-trivial cost: a sandbox costs CPU, while a recorder costs cluster infrastructure plus an engineer who knows how to operate it.

\section{Problem Statement and Research Gap}

This thesis addresses three concrete gaps that emerge from the literature and from the state of the art surveyed in Chapter~\ref{ch:litreview}.

\textbf{Gap 1: missing comparative study.} No recent work compares a properly hardened traditional runtime against sandboxed alternatives. Existing comparisons~\cite{wang2022,volpert2024} use \emph{default} \texttt{runc} configurations, leaving the security--performance trade-off of an opinionated, hardened \texttt{runc} unmeasured.

\textbf{Gap 2: developer-grade automation is missing.} Current automation either lives in a pipeline (SPO, Inspektor Gadget recordings, Tetragon) or in research prototypes that target individual mechanisms~\cite{lopes2020,canella2021,mattetti2020}. A solo developer or a small startup that wants confinement comparable to what a DevSecOps team would produce in a CI matrix has no equivalent tool that operates at the runtime layer itself.

\textbf{Gap 3: limited integration across mechanisms.} Most prior automated approaches~\cite{lopes2020,canella2021} target seccomp alone. AppArmor, capability narrowing, and the orchestration of all three under one workflow remain fragmented across separate utilities and bundles.

\section{Research Questions}

\begin{enumerate}
    \item Can the runtime itself, without infrastructure changes, generate seccomp, AppArmor, and capability profiles whose security level approaches what a CI-driven recorder produces?
    \item How should a runtime expose this capability so that it remains a one-flag operation for a developer, while still satisfying the constraints CI-side recorders solve (child-process coverage, profile relaxation during the recording, non-root requirement)?
    \item What is the resulting trade-off between the hardened traditional runtime, the same runtime in default mode, and sandboxed alternatives (gVisor, Kata)?
\end{enumerate}

\section{Proposed Approach}

The proposed runtime is a fork of \texttt{runc} that adds a single flag, \texttt{-{}-security-scan}. When the flag is present:

\begin{itemize}
    \item All host-supplied confinement is relaxed in memory for the run (seccomp cleared, SELinux label cleared, \texttt{NoNewPrivileges} disabled, all known capabilities granted, AppArmor swapped for a complain-mode profile). The on-disk \texttt{config.json} is not mutated.
    \item Syscalls are recorded by \texttt{oci-seccomp-bpf-hook} and emitted as \texttt{generated/seccomp.json}.
    \item Capabilities are traced cgroup-wide by \texttt{capable-bpfcc} with a \texttt{bpftool}-pinned cgroup map, so child processes are observed (not only the init PID).
    \item AppArmor learns through a complain-mode profile loaded by a \texttt{createRuntime} hook and unloaded by a \texttt{poststop} hook.
    \item All hooks are sub-commands of the runtime binary itself (\texttt{runc scan-aa-load}, \texttt{runc scan-cap-trace-start}, etc.). The OCI bundle never has to embed executable artefacts.
    \item On finalisation, the runtime narrows --- never widens --- \texttt{process.capabilities} in \texttt{config.json} to the observed set.
\end{itemize}

A second binary mode, the \emph{enforcement} mode, simply runs the workload with the generated profiles applied. The implementation also revisits the default capability set and ships two opt-in flags (\texttt{-{}-default-capabilities} for the upstream three-cap set, \texttt{-{}-default-capabilities-docker} for the historical fourteen-cap set), so that the hardened default does not break existing bundles.

\section{Experimental Evaluation}

The runtime is compared against four configurations:

\begin{itemize}
    \item \textbf{\texttt{runc} (default):} Standard OCI runtime with stock policies.
    \item \textbf{Proposed runtime:} \texttt{--security-scan} for the recording run, then enforcement mode for the steady-state runs.
    \item \textbf{gVisor:} User-space kernel sandbox.
    \item \textbf{Kata Containers:} Microvm-based isolation.
\end{itemize}

The primary workload is an intentionally vulnerable Flask application in the \texttt{runc-hardened-test} laboratory, used to measure post-exploitation containment after command injection. Reported metrics cover profile generation quality, containment effectiveness, startup latency, and CPU and memory overhead. Results are reported in Chapter~\ref{ch:experiments} and discussed in Chapter~\ref{ch:discussion}.

\section{Expected Contribution}

\begin{enumerate}
    \item The first published direct comparison between an opinionated, automatically hardened \texttt{runc} and the dominant sandboxed runtimes.
    \item A practical, developer-grade runtime that produces seccomp, AppArmor, and capability profiles from a single CLI flag, with no cluster, recorder, or operator prerequisites.
    \item Evidence-based guidance for organisations choosing between runtime-side automation, cluster-side recorders, and full sandboxing.
    \item Reproducible evaluation harnesses (\texttt{runc-hardened-test} for security, \texttt{performance-evaluation} for benchmarks) that document how the claims in this thesis can be replayed on a stock Linux host.
\end{enumerate}

\section{Thesis Organization}

Chapter~\ref{ch:litreview} reviews the relevant literature and the competitive landscape from manual authoring to full sandboxing. Chapter~\ref{ch:design} describes the design goals, the threat model, and the scanner architecture. Chapter~\ref{ch:impl} covers the implementation. Chapter~\ref{ch:experiments} reports the experimental setup and results. Chapter~\ref{ch:discussion} interprets those results and states threats to validity. Chapter~\ref{ch:conclusion} concludes and outlines future work.
